---
title: "'The More the Trials, the Stronger the Effect'. Fact or Myth? Evidence from Classical Experimental Psychology Paradigms"
subtitle: "Supplementary Materials"
toc: true
format:
  pdf:
    documentclass: article
    papersize: a4
  html: default
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(latex2exp)
library(stringr)
library(patchwork)

source(here("paper", "_paper-utils.R"), local = knitr::knit_global())
set_paper_theme()

knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  fig.width = 7.2,
  fig.height = 4.8,
  dpi = 300,
  out.width = "100%"
)
```

```{r}
#| label: load-supplementary-results
#| cache: true

res <- readRDS(
  here("objects", "task_cum_log.rds")
)

res_by_subj <- readRDS(
  here("objects", "task_cum_by_subj.rds")
)

res_plot <- res |>
  select(
    task,
    trial,
    params_log,
    emmeans_log
  ) |>
  rename(
    params = params_log,
    emmeans = emmeans_log
  )
```

\newpage

# Log-normal model

In this section, we include the results of the same cumulative models presented in the main paper, but using a log transformation of reaction times. We decided to present raw reaction times in the main text to improve the interpretability of the model parameters, particularly the variance decomposition.

```{r}
#| label: fig-log-cum-simon

plot_cumulative_task(
  res_plot,
  "simon",
  "Simon"
)
```

```{r}
#| label: fig-log-cum-snarc

plot_cumulative_task(
  res_plot,
  "snarc",
  "SNARC"
)
```

```{r}
#| label: fig-log-cum-tswitch

plot_cumulative_task(
  res_plot,
  "tswitch",
  "Task-switching"
)
```

# By-subject cumulative effect

We also computed a by-subject cumulative effect by applying the same model separately to each participant. Clearly, we only have the estimation of $\beta_0$, $\beta_1$ and $\sigma_{E}$.

```{r}
#| label: define-by-subject-plot

plot_by_subject <- function(
    data,
    task_id) {
  data |>
    unnest(fit) |>
    filter(
      task == task_id
    ) |>
    ggplot(
      aes(
        x = trial,
        y = estimate,
        group = id
      )
    ) +
    facet_wrap(
      ~term,
      ncol = 1,
      scales = "free",
      labeller = labeller(
        term = term_facet_labeller
      )
    ) +
    geom_line(
      alpha = 0.1
    ) +
    labs(
      title = unname(
        task_labels[task_id]
      ),
      x = "Cumulative trial",
      y = NULL
    )
}
```

```{r}
#| label: fig-by-subject-simon
#| fig-height: 10

plot_by_subject(
  res_by_subj,
  "simon"
)
```

```{r}
#| label: fig-by-subject-snarc
#| fig-height: 10

plot_by_subject(
  res_by_subj,
  "snarc"
)
```

```{r}
#| label: fig-by-subject-tswitch
#| fig-height: 10

plot_by_subject(
  res_by_subj,
  "tswitch"
)
```
