---
title: "'The More the Trials, the Stronger the Effect'. Fact or Myth? Evidence from Classical Experimental Psychology Paradigms"
subtitle: "Supplementary Materials"
toc: true
format:
  pdf: default
  html: default
---

```{r}
#| label: setup
#| include: false

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(latex2exp)
library(stringr)
library(patchwork)

knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  fig.width = 7.2,
  fig.height = 4.8,
  dpi = 300,
  out.width = "100%"
)

plot_base_size <- 11

plot_lwd <- 0.65
plot_lwd_strong <- 0.80
plot_lwd_ref <- 0.45

point_size <- 1.25
point_alpha <- 0.35
line_alpha <- 0.85
ribbon_alpha <- 0.22

common_theme <- function(base_size = plot_base_size) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(size = base_size),

      axis.title = element_text(
        face = "bold",
        size = base_size + 1
      ),
      axis.text = element_text(size = base_size - 1),

      strip.text = element_text(
        face = "bold",
        size = base_size
      ),
      strip.background = element_rect(
        fill = "grey92",
        color = NA
      ),

      plot.title = element_text(
        face = "bold",
        size = base_size + 2,
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      plot.subtitle = element_text(
        size = base_size,
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      plot.caption = element_text(
        size = base_size - 2,
        hjust = 0
      ),

      legend.title = element_blank(),
      legend.text = element_text(size = base_size - 1),
      legend.position = "bottom",

      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.25),
      panel.spacing = grid::unit(1.1, "lines"),

      plot.margin = margin(6, 6, 6, 6)
    )
}

theme_set(common_theme())

update_geom_defaults(
  "line",
  list(linewidth = plot_lwd, alpha = line_alpha)
)
update_geom_defaults(
  "smooth",
  list(linewidth = plot_lwd_strong)
)
update_geom_defaults(
  "point",
  list(size = point_size, alpha = point_alpha)
)
update_geom_defaults(
  "ribbon",
  list(alpha = ribbon_alpha, colour = NA)
)
update_geom_defaults(
  "boxplot",
  list(linewidth = plot_lwd_ref, outlier.size = 1.1)
)

geom_paper_line <- function(
    ...,
    linewidth = plot_lwd_strong,
    alpha = line_alpha) {
  geom_line(
    ...,
    linewidth = linewidth,
    alpha = alpha
  )
}

geom_paper_ribbon <- function(
    ...,
    alpha = ribbon_alpha) {
  geom_ribbon(
    ...,
    alpha = alpha,
    colour = NA
  )
}

geom_paper_hline <- function(yintercept = 0, ...) {
  geom_hline(
    yintercept = yintercept,
    linewidth = plot_lwd_ref,
    linetype = "dashed",
    ...
  )
}

geom_paper_point <- function(
    ...,
    size = point_size,
    alpha = point_alpha) {
  geom_point(
    ...,
    size = size,
    alpha = alpha
  )
}

task_labels <- c(
  simon = "Simon",
  snarc = "SNARC",
  tswitch = "Task-switching"
)

condition_labels <- function(task, cond) {
  case_when(
    task == "tswitch" & cond == "i" ~ "Switch",
    task == "tswitch" & cond == "c" ~ "Non-switch",
    cond == "i" ~ "Incongruent",
    cond == "c" ~ "Congruent"
  )
}

term_titles <- c(
  "(Intercept)" = TeX("$\\beta_0$", output = "character"),
  "condi" = TeX("$\\beta_1$", output = "character"),
  "cond1" = TeX("$\\beta_1$", output = "character"),
  "sd__(Intercept)" = TeX("$\\sigma_P$", output = "character"),
  "sd__cond1" = TeX(
    "$\\sigma_{P \\times C}$",
    output = "character"
  ),
  "sd__condi" = TeX(
    "$\\sigma_{P \\times C}$",
    output = "character"
  ),
  "sd__Observation" = TeX(
    "$\\sigma_E$",
    output = "character"
  ),
  "sigma" = TeX(
    "$\\sigma_E$",
    output = "character"
  )
)

term_facet_labeller <- as_labeller(
  term_titles,
  default = label_parsed
)
```

```{r}
#| label: data
#| cache: true

res <- readRDS(
  here("objects", "task_cum_log.rds")
)

res_by_subj <- readRDS(
  here("objects", "task_cum_by_subj.rds")
)

res_plot <- res |>
  select(
    task,
    trial,
    params_log,
    emmeans_log
  ) |>
  rename(
    params = params_log,
    emmeans = emmeans_log
  )
```

\newpage

# Log-normal model

In this section, we include the results of the same cumulative models presented in the main paper, but using a log transformation of reaction times. We decided to present raw reaction times in the main text to improve the interpretability of the model parameters, particularly the variance decomposition.

```{r}
cum_plot <- function(
    data,
    x_limits = NULL,
    title = NULL) {
  data <- data |>
    mutate(
      conf.low = if_else(
        is.na(conf.low),
        estimate,
        conf.low
      ),
      conf.high = if_else(
        is.na(conf.high),
        estimate,
        conf.high
      )
    )

  parsed_title <- if (is.null(title)) {
    NULL
  } else {
    parse(text = title)[[1]]
  }

  ggplot(
    data,
    aes(
      x = trial,
      y = estimate
    )
  ) +
    geom_paper_ribbon(
      aes(
        ymin = conf.low,
        ymax = conf.high
      )
    ) +
    geom_paper_line() +
    scale_x_continuous(
      limits = x_limits
    ) +
    labs(
      title = parsed_title,
      x = "Cumulative trial",
      y = NULL
    )
}

emm_limits <- function(data) {
  lower_name <- intersect(
    c(
      "asymp.LCL",
      "lower.CL",
      "conf.low"
    ),
    names(data)
  )[1]

  upper_name <- intersect(
    c(
      "asymp.UCL",
      "upper.CL",
      "conf.high"
    ),
    names(data)
  )[1]

  list(
    lower = lower_name,
    upper = upper_name
  )
}

plot_cumulative_task <- function(
    data,
    task_id,
    task_title) {
  cum_dat <- data |>
    filter(task == task_id) |>
    unnest(params) |>
    filter(
      !str_detect(term, "^cor__")
    )

  x_limits <- range(
    cum_dat$trial,
    na.rm = TRUE
  )

  emm_dat <- data |>
    filter(task == task_id) |>
    unnest(emmeans) |>
    mutate(
      cond = condition_labels(
        task,
        cond
      ),
      cond = factor(
        cond,
        levels = c(
          "Congruent",
          "Incongruent",
          "Non-switch",
          "Switch"
        )
      )
    ) |>
    filter(
      contrast != "c - i"
    )

  ci_names <- emm_limits(emm_dat)

  p_emm <- ggplot(
    emm_dat,
    aes(
      x = trial,
      y = emmean,
      color = cond,
      fill = cond
    )
  ) +
    geom_paper_ribbon(
      aes(
        ymin = .data[[ci_names$lower]],
        ymax = .data[[ci_names$upper]]
      ),
      alpha = ribbon_alpha
    ) +
    geom_paper_line() +
    scale_x_continuous(
      limits = x_limits
    ) +
    labs(
      title = NULL,
      x = "Cumulative trial",
      y = NULL
    ) +
    theme(
      legend.position = "bottom",
      legend.background = element_rect(
        fill = scales::alpha(
          "white",
          0.75
        ),
        color = NA
      ),
      legend.key = element_rect(
        fill = scales::alpha(
          "white",
          0.75
        ),
        color = NA
      )
    )

  make_term_plot <- function(term_candidates) {
    term_present <- intersect(
      term_candidates,
      unique(cum_dat$term)
    )[1]

    if (is.na(term_present)) {
      return(plot_spacer())
    }

    title <- term_titles[[term_present]]

    if (is.null(title)) {
      title <- term_present
    }

    cum_dat |>
      filter(
        term == term_present
      ) |>
      cum_plot(
        x_limits = x_limits,
        title = title
      )
  }

  plot_list <- list(
    make_term_plot("(Intercept)"),
    make_term_plot(c("condi", "cond1")),
    p_emm,
    make_term_plot("sd__(Intercept)"),
    make_term_plot(
      c("sd__condi", "sd__cond1")
    ),
    make_term_plot(
      c("sd__Observation", "sigma")
    )
  )

  wrap_plots(
    plot_list,
    ncol = 3
  ) +
    plot_layout(
      axes = "collect",
      axis_titles = "collect"
    ) +
    plot_annotation(
      title = task_title
    )
}
```

```{r}
plot_cumulative_task(
  res_plot,
  "simon",
  "Simon"
)
```

```{r}
plot_cumulative_task(
  res_plot,
  "snarc",
  "SNARC"
)
```

```{r}
plot_cumulative_task(
  res_plot,
  "tswitch",
  "Task-switching"
)
```

# By-subject cumulative effect

We also computed a by-subject cumulative effect by applying the same model separately to each participant. Clearly, we only have the estimation of $\beta_0$, $\beta_1$ and $\sigma_{E}$.

```{r}
plot_by_subject <- function(
    data,
    task_id) {
  data |>
    unnest(fit) |>
    filter(
      task == task_id
    ) |>
    ggplot(
      aes(
        x = trial,
        y = estimate,
        group = id
      )
    ) +
    facet_wrap(
      ~term,
      ncol = 1,
      scales = "free",
      labeller = labeller(
        term = term_facet_labeller
      )
    ) +
    geom_line(
      alpha = 0.1
    ) +
    labs(
      title = unname(
        task_labels[task_id]
      ),
      x = "Cumulative trial",
      y = NULL
    )
}
```

```{r}
#| fig-height: 10
plot_by_subject(
  res_by_subj,
  "simon"
)
```

```{r}
#| fig-height: 10
plot_by_subject(
  res_by_subj,
  "snarc"
)
```

```{r}
#| fig-height: 10
plot_by_subject(
  res_by_subj,
  "tswitch"
)
```
